---
title: "Geographic data processing with R"
subtitle: "A practical introduction to data manipulation and visualization"
author: "Damiano Baldan"
date: "2026-02-25"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why R for geographic data processing?

-   R is a powerful and flexible programming language for data analysis and visualization.
-   R has a rich ecosystem of packages for working with geographic data (e.g. `sf` for vector data and `terra` for raster data)
-   R allows for reproducible research and data processing workflows, making it easier to share and collaborate on projects.
-   R has a large and active community of users and developers, providing support and resources for learning and troubleshooting.
-   R allows for easier integration with downstream analyses (e.g. statistical modeling, machine learning, etc.) and visualization (e.g. `ggplot2`, etc.)

## What we will cover in this presentation

-   Retrieval of geographic data from various sources (e.g. HydroSHEDS, HydroATLAS, HydroRIVERS)
-   Processing of geographic data using R packages (e.g. `sf`, `terra`, `dplyr`, etc.)
-   Basic spatial operations (e.g. spatial joins, buffering, etc.)
-   Visualization of geographic data using R packages (e.g. `ggplot2`, etc.)

## Which packages we will use

-   `sf` for vector data manipulation: sf represents spatial data in a simple, data.frame-like format
-   `terra` for raster data manipulation: it's the successor of the `raster` package and has a lot of useful functions for raster data
-   `dplyr` for data manipulation: helps building pipelines for easy-to-read data manipulation
-   `ggplot2` for data visualization: a bit painful, but easily integrates `sf` and `terra` objects for plotting

-   `rnaturalearth` for retrieving natural earth data: provides easy access to natural earth vector data (e.g. country borders, etc.)

```{r libraries, echo = TRUE}
library(tidyverse)
library(sf)
library(terra)
library(rnaturalearth)
```

## The HydroSHEDS dataset

-   [HydroSHEDS](https://www.hydrosheds.org/) (Hydrological data and maps based on SHuttle Elevation Derivatives at multiple Scales) is a global dataset of hydrological information derived from satellite data.
-   It provides high-resolution data on river networks, watershed boundaries, and other hydrological features
-   Data product [HydroRIVERS](https://www.hydrosheds.org/products/hydrorivers): river network data, including flow direction and flow accumulation
-   Data product [HydroBASINS](https://www.hydrosheds.org/products/hydrobasins): outlines of river basins at multiple scales (from level 1 to level 12, with increasing detail)
-   Data product [HydroATLAS](https://www.hydrosheds.org/hydroatlas): a collection of attributes for river basins, including physical, climatic, and anthropogenic characteristics

## HydroBASINS dataset

-   Different levels of details (1 to 12) with increasing resolution and number of basins
-   The dataset can be downloaded as shapefile and has to be processed locally
-   Here the data are already cropped for the Ebro river basin (Spain) at levels 5 and 10
-   Let's load in the environment the level 5 shapefile and check its content

## HydroBASINS dataset

```{r hydrosheds read, echo = TRUE}
hydrosheds_lev5 <- "data/HydroBASINS_sample/sample__hybas_eu_lev05_v1c.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()

```

-   The pipeline operator `%>%` concatenates functions
-   `st_read()` reads the shapefile and creates an sf object
-   `st_make_valid()` ensures that the geometries are valid (e.g. no self-loops)

## HydroBASINS dataset

We check the shapefile:

```{r hydrosheds class, echo = TRUE}
class(hydrosheds_lev5)

```

-   It is an object of class `sf`
-   It is also an object of class `data.frame`: all data.frame functions work :)

## HydroBASINS dataset

We check the shapefile content:

```{r hydrosheds head, echo = TRUE}
head(hydrosheds_lev5)

```

-   Is basically a data.frame with attributes and a geometry column
-   Field HYBAS_ID is the unique identifier of each basin, and SUB_AREA is the area of the basin

## HydroBASINS dataset

We plot the shapefile using `ggplot` and color-coding the attribute SUB_AREA:

```{r hydrosheds plot1, echo=TRUE}
plot <- ggplot() + 
  geom_sf(data = hydrosheds_lev5, aes(fill = SUB_AREA))

```


## HydroBASINS dataset

We plot the shapefile using `ggplot` and color-coding the attribute SUB_AREA:

```{r hydrosheds plot1 setup, echo=TRUE}
plot <- ggplot() + 
  geom_sf(data = hydrosheds_lev5, aes(fill = SUB_AREA))

```

## HydroBASINS dataset

```{r hydrosheds plot1 plot, echo=FALSE}
plot

```

## HydroBASINS dataset

We try to make the plot a bit nicer.

```{r hydrosheds plot2 setup1, echo=TRUE}
crs_data <- sf::st_crs(hydrosheds_lev5) 

world <- rnaturalearth::ne_countries(scale="medium", returnclass="sf") %>%
  sf::st_transform(crs = crs_data ) %>%
  sf::st_as_sf(coords=c("Long", "Lat"), crs = crs_data )

bbox <- sf::st_bbox(hydrosheds_lev5) %>%  
  st_as_sfc() %>%   
  st_buffer(1000 * 1000)

neighbors_cropped <- sf::st_crop(world, bbox) %>% dplyr::select(tlc, geometry)


```

## HydroBASINS dataset

We try to make the plot a bit nicer. First we add the HydroBASINS shapefile, then we add the background with the countries borders. Note that we set the limits of the plot manually.

```{r hydrosheds plot2 setup2, echo=TRUE}
plot <- ggplot() + 
  theme_bw() + 
  geom_sf(data = hydrosheds_lev5, aes(fill = SUB_AREA)) + 
  geom_sf(data = neighbors_cropped, 
          fill = "gray30", 
          color = "gray80", 
          alpha = 0.5, linewidth = 0.3) + 
  xlim(-10, 10) + 
  ylim(35, 50)


```

## HydroBASINS dataset

```{r hydrosheds plot2 plot, echo=FALSE}
plot 

```



























