---
title: "Analysis of fragmented river systems"
subtitle: "Lecture 1: Geographic data processing with R"
author: "Damiano Baldan"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:
    widescreen: true
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, '1_data_processing.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why R for geographic data processing?

-   R is a powerful and flexible programming language for data analysis and visualization.
-   R has a rich ecosystem of packages for working with geographic data (e.g. `sf` for vector data and `terra` for raster data)
-   R allows for reproducible research and data processing workflows, making it easier to share and collaborate on projects.
-   R has a large and active community of users and developers, providing support and resources for learning and troubleshooting.
-   R allows for easier integration with downstream analyses (e.g. statistical modeling, machine learning, etc.) and visualization (e.g. `ggplot2`, etc.)

## What we will cover in this presentation

-   Retrieval of geographic data from various sources (e.g. HydroSHEDS, HydroATLAS, HydroRIVERS);
-   Processing of geographic data using R packages (e.g. `sf`, `terra`, `dplyr`, etc.);
-   Basic spatial operations (e.g. spatial joins, etc.);
-   Visualization of geographic data using R packages (e.g. `ggplot2`, etc.).

## Which packages we will use

-   `sf` for vector data manipulation: sf represents spatial data in a simple, data.frame-like format.
-   `terra` for raster data manipulation: it's the successor of the `raster` package and has a lot of useful functions for raster data.
-   `dplyr` for data manipulation: helps building pipelines for easy-to-read data manipulation (loaded with the `tidyverse`).
-   `ggplot2` for data visualization: a bit painful, but easily integrates `sf` and `terra` objects for plotting (loaded with the `tidyverse`).
-   `rnaturalearth` for retrieving natural earth data: provides easy access to natural earth vector data (e.g. country borders, etc.).
-   `tidyterra` for integrating `terra` objects with the `tidyverse`.
-   `elevatr` for retrieving elevation data: provides easy access to elevation data from various sources (e.g. AWS Terrain Tiles, etc.).

## Which packages we will use

```{r libraries, echo = TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(terra)
library(rnaturalearth)
library(tidyterra)
library(viridis)
library(elevatr)
```

## The HydroSHEDS dataset

-   [HydroSHEDS](https://www.hydrosheds.org/) (Hydrological data and maps based on SHuttle Elevation Derivatives at multiple Scales) is a global dataset of hydrological information derived from satellite data.
-   It provides high-resolution data on river networks, watershed boundaries, and other hydrological features
-   Data product [HydroRIVERS](https://www.hydrosheds.org/products/hydrorivers): river network data, including flow direction and flow accumulation
-   Data product [HydroBASINS](https://www.hydrosheds.org/products/hydrobasins): outlines of river basins at multiple scales (from level 1 to level 12, with increasing detail)
-   Data product [HydroATLAS](https://www.hydrosheds.org/hydroatlas): a collection of attributes for river basins, including physical, climatic, and anthropogenic characteristics

# The HydroBASINS dataset

## HydroBASINS dataset

We start with [HydroBASINS](https://www.hydrosheds.org/products/hydrobasins) data.

-   Different levels of details (1 to 12) with increasing resolution and number of basins
-   The dataset can be downloaded as shapefile and has to be processed locally
-   Here the data are already cropped for the Ebro river basin (Spain) at levels 5 and 10
-   Let's load in the environment the level 5 shapefile and check its content
-   All the data we will work on are available in the  `data/` folder

## HydroBASINS dataset

```{r hydrobasins read, echo = TRUE}
hydrosheds_lev5 <- "data/HydroBASINS_sample/sample__hybas_eu_lev05_v1c.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()

```

-   The pipeline operator `%>%` concatenates functions
-   `st_read()` reads the shapefile and creates an sf object
-   `st_make_valid()` ensures that the geometries are valid (e.g. no self-loops)

## HydroBASINS dataset

We check the shapefile:

```{r hydrobasins class, echo = TRUE}
class(hydrosheds_lev5)

```

-   It is an object of class `sf`
-   It is also an object of class `data.frame`: all data.frame functions work :)
-   The object is basically a data.frame with some other attributes and a geometry column

## HydroBASINS dataset

We check the shapefile content:

```{r hydrobasins head, echo = TRUE}
head(hydrosheds_lev5)

```

-   the field HYBAS_ID is the unique identifier of each basin.
-   Use HYBAS_ID to filter full dataset to area of interest (in this case, HYBAS_ID == 2050017150, check e.g. with QGIS).

## HydroBASINS dataset

We plot the shapefile using `ggplot` and color-coding the attribute SUB_AREA:

```{r hydrobasins plot1, echo=TRUE}
plot <- ggplot() + 
  geom_sf(data = hydrosheds_lev5, aes(fill = SUB_AREA))

```


## HydroBASINS dataset

```{r hydrobasins plot1 plot, echo=FALSE}
plot

```

## HydroBASINS dataset

We try to make the plot a bit nicer. We use `rnaturalearth::ne_countries` to retrieve the world shapefile; `st_bbox()` to get the bounding box of the shapefile; `st_crop()` to crop the world shapefile to the bounding box.

```{r hydrobasins plot2 setup1, echo=TRUE, comment = NA}
world <- rnaturalearth::ne_countries(scale="medium", returnclass="sf") 

bbox <- hydrosheds_lev5 %>% 
  sf::st_bbox() %>%  
  sf::st_as_sfc() %>%   
  sf::st_buffer(1000 * 1000)

neighbors_cropped <- world %>% 
  sf::st_crop(bbox) %>% 
  dplyr::select(tlc, geometry)

```

## HydroBASINS dataset

We try to make the plot a bit nicer. First we add the the background with the countries borders, then the HydroBASINS shapefile. Note that we set the limits of the plot manually.

```{r hydrobasins plot2 setup2, echo=TRUE}
plot <- ggplot() + 
  theme_bw() + 
  geom_sf(data = neighbors_cropped, 
          fill = "gray30", 
          color = "gray80", 
          alpha = 0.5, linewidth = 0.3) + 
  geom_sf(data = hydrosheds_lev5, aes(fill = SUB_AREA)) + 
  xlim(-10, 10) + 
  ylim(35, 50)


```

## HydroBASINS dataset

```{r hydrobasins plot2 plot, echo=FALSE}
plot 

```

## HydroBASINS dataset level 10

Now we move forward and check the level 10, that has higher resolution.

```{r hydrobasins9 read, echo = TRUE}
hydrosheds_lev10 <- "data/HydroBASINS_sample/sample__hybas_eu_lev10_v1c.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()

```

## HydroBASINS dataset level 10

We check the shapefile content:

```{r hydrobasins10 head, echo = TRUE}
head(hydrosheds_lev10)

```

-   Field HYBAS_ID is the unique identifier of each basin
-   Use MAIN_BAS to filter full dataset to area of interest (in this case, MAIN_BAS == 2100017150, check with e.g. QGIS)

## HydroBASINS dataset level 10

We plot the shapefile using `ggplot` and color-coding the attribute SUB_AREA. We use `scale_fill_viridis()` to set the color palette

```{r hydrobasins10 plot1, echo=TRUE}
plot <- ggplot() + 
  theme_bw() + 
  geom_sf(data = neighbors_cropped, 
          fill = "gray30", 
          color = "gray80", 
          alpha = 0.5, linewidth = 0.3) +
  geom_sf(data = hydrosheds_lev10, aes(fill = SUB_AREA)) + 
  scale_fill_viridis() + 
  xlim(-5, 3) + 
  ylim(40, 44)

```

## HydroBASINS dataset level 10

```{r hydrobasins10 plot1 plot, echo=FALSE}
plot

```


# Extracting data from rasters

## Extracting data from raster sources

-   HydroBASINS does not contain much information on the environmental conditions of the basins.
-   We can add this information with some spatial operations.
-   For now we focus on elevation and slope.
-   We download the elevation data using the `elevatr` package, that retrieves elevation data from the [AWS Terrain Tiles dataset](https://registry.opendata.aws/terrain-tiles/).

```{r elevation , echo=TRUE, message=FALSE, warning=FALSE}
elevation <- hydrosheds_lev5 %>% 
  elevatr::get_elev_raster(z = 8) %>% 
  terra::rast()

names(elevation) <- "elevation"

```

## Extracting data from raster sources

We check the object elevation.

```{r elevation check, echo=TRUE, message=FALSE, warning=FALSE}
elevation
```

## Extracting data from raster sources

The downloaded data are in a raster format and we can plot them directly.

```{r elevation plot , echo=TRUE, message=FALSE, warning=FALSE}
plot(elevation)
```

## Extracting data from raster sources

We can also produce a plot with ggplot where we superimpose other informations. Note that `geom_sf` does not have any aestetics (`aes()`), this is because we do not want things to change based on the attributes of the shapefile, but we just want to use it as a background.

```{r elevation plot2 , echo=TRUE, message=FALSE, warning=FALSE}
plot <- ggplot() + 
  theme_bw() + 
  tidyterra::geom_spatraster(data = elevation) + 
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) +
  geom_sf(data = hydrosheds_lev10, fill = NA, color = "gray90", alpha = 0.5) + 
  scale_fill_viridis(name = "Elevation (m)")
```

## Extracting data from raster sources

```{r elevation plot2 plot , echo=FALSE, message=FALSE, warning=FALSE}
plot 
```

## Extracting data from raster sources

Based on the elevation data, we use the `terrain` function to calculate the slope and we plot it.

```{r slope plot , echo=TRUE, message=FALSE, warning=FALSE}
slope <- elevation %>% 
  terra::terrain(v = "slope", unit = "degrees")

names(slope) <- "slope"

plot <- ggplot() + 
  theme_bw() + 
  tidyterra::geom_spatraster(data = slope) + 
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = hydrosheds_lev10, fill = NA, color = "gray90", alpha = 0.5) + 
  scale_fill_viridis(name = "Slope (deg.)")
```

## Extracting data from raster sources

```{r slope plot plot , echo=FALSE, message=FALSE, warning=FALSE}
plot 
```

## Extracting data from raster sources

Now we would like to extract the information on the elevation and the slope and assign to each subcatchment the related information. We can do this with the `extract` function of the `terra` package, that extracts the values of a raster based on the geometries of a shapefile. Then, the information is added to the shapefile with `dplyr::mutate`. We use the `fun = mean` argument to get the average elevation and slope of each subcatchment.

```{r extract elevation and slope, echo=TRUE, message=FALSE, warning=FALSE}

elevation_extracted <- elevation %>% 
  terra::extract(hydrosheds_lev10, fun = mean, na.rm = TRUE)

slope_extracted <- slope %>% 
  terra::extract(hydrosheds_lev10, fun = mean, na.rm = TRUE)

hydrosheds_lev10 <- hydrosheds_lev10 %>% 
  dplyr::mutate( 
    elevation = elevation_extracted$elevation, 
    slope = slope_extracted$slope)

```

## Extracting data from raster sources

We can add elevation and slope to the plot.

```{r elevation and slope add to plot, echo=TRUE, message=FALSE, warning=FALSE}
plot_elev <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = hydrosheds_lev10, aes(fill = elevation)) + 
  scale_fill_viridis(name = "Elevation (m)")

plot_slope <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = hydrosheds_lev10, aes(fill = slope)) + 
  scale_fill_viridis(name = "Slope (degrees)")

```

## Extracting data from raster sources

```{r elevation plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_elev
```

## Extracting data from raster sources

```{r slope plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_slope
```

# The HydroRIVERS dataset

## HydroRIVERS dataset

Now we focus on the [HydroRIVERS](https://www.hydrosheds.org/products/hydrorivers) dataset. The data are already cropped for the Ebro river basin (Spain). We see that the structure is similar to that of HydroSHEDS, except here the geometry type is LINESTRING instead of POLYGON.

```{r read hydrorivers, echo=TRUE, message=FALSE, warning=FALSE}
hydrorivers <- "data/HydroRIVERS_sample/sample__HydroRIVERS_v10_eu.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()
```

## HydroRIVERS dataset

HydroRIVERS has several interesting attributes, such as the river order (ORD_STRA), the length of the river (LENGTH_KM), the average upland area (UPLAND_SKM), etc.

```{r check hydrorivers, echo=TRUE, message=FALSE, warning=FALSE}
head(hydrorivers)
```

## HydroRIVERS dataset

`ggplot` can be used to plot hydrorivers. Here we color-code the river segments based on the LENGTH_KM field, and transparency-code based on stream order (ORD_STRA).

```{r hydrorivers plot , echo=TRUE, message=FALSE, warning=FALSE}
plot <- ggplot() + 
  theme_bw() + 
  geom_sf(data = hydrosheds_lev5, fill = "white", color = "gray50", alpha = 0.5) +
  geom_sf(data = hydrorivers, aes(color = LENGTH_KM, alpha = ORD_STRA)) + 
  scale_color_viridis(name = "Length (km)") + 
  scale_alpha_continuous(name = "Stream order")
```

## HydroRIVERS dataset

```{r hydrorivers plot plot, echo=FALSE, message=FALSE, warning=FALSE}
plot
```

## HydroRIVERS dataset

The hydrorivers object is quite dense (~5000 river segments), so the plot is quite busy. In many cases, it might be useful to filter out low-order segments.


```{r hydrorivers filter, echo=TRUE, message=FALSE, warning=FALSE}
nrow(hydrorivers)

hydrorivers_filtered <- hydrorivers %>% 
  dplyr::filter(ORD_STRA >= 3)

nrow(hydrorivers_filtered)
```

## HydroRIVERS dataset

```{r hydrorivers filtered plot, echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  theme_bw() + 
  geom_sf(data = hydrosheds_lev5, fill = "white", color = "gray50", alpha = 0.5) +
  geom_sf(data = hydrorivers_filtered, aes(color = LENGTH_KM, alpha = ORD_STRA)) + 
  scale_color_viridis(name = "Length (km)") + 
  scale_alpha_continuous(name = "Stream order")

```

## HydroRIVERS dataset

The elevation and slope raster data can be extracted also for the river segments using the `extract` function as before.The function extracts the raster values underlying each segment and then calculates their mean value.

```{r elevation slope extract hydrorivers, echo=TRUE, message=FALSE, warning=FALSE}
elevation_extracted <- elevation %>% 
  terra::extract(hydrorivers, fun = mean, na.rm = TRUE)

slope_extracted <- slope %>% 
  terra::extract(hydrorivers, fun = mean, na.rm = TRUE)

hydrorivers <- hydrorivers %>% 
  dplyr::mutate( elevation = elevation_extracted$elevation, 
                 slope = slope_extracted$slope)
```

## HydroRIVERS dataset

The extracted values can be plotted. Not that now the `scale_color_viridis()` is used to color-code the river segments based on the elevation and slope values.

```{r elevation slope hydrorivers plot, echo=TRUE, message=FALSE, warning=FALSE}
plot_elev <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = hydrorivers, aes(color = elevation)) + 
  scale_color_viridis(name = "Elevation (m)")

plot_slope <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = hydrorivers, aes(color = slope)) + 
  scale_color_viridis(name = "Slope (deg.)")
```

## HydroRIVERS dataset

```{r elevation hydrorivers plot plot , echo=FALSE, message=FALSE, warning=FALSE}
plot_elev
```

## HydroRIVERS dataset

```{r slope hydrorivers plot plot , echo=FALSE, message=FALSE, warning=FALSE}
plot_slope
```


# The HydroATLAS dataset

## HydroATLAS dataset

Finally, we load the [HydroATLAS](https://www.hydrosheds.org/hydroatlas) dataset, that contains a collection of attributes for river basins, including physical, climatic, and anthropogenic characteristics. 

The dataset contains both data aggregated at the subbasin level and data aggregated for river segments. The data are already cropped for the Ebro river basin.

To retain the data for the Ebro catchment:

-   For BasinATLAS, the original data can be filtered based on the fields MAIN_BAS == 2100017150 (check in e.g. QGIS)
-   For RiverATLAS, the original data can be filtered based on the field MAIN_RIV == 20598415 (check in e.g. QGIS)


## HydroATLAS dataset

RiverATLAS has the same structure of HydroRIVERS, but has more fields. Data from several sources were used to annotate the river shapefile. The sources are listed in the documentation.

```{r read riveratlas , echo=TRUE, message=FALSE, warning=FALSE}
riveratlas <- "data/HydroATLAS_sample/sample__RiverATLAS_v10_eu.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()
```


## HydroATLAS dataset

BasinATLAS has the same structure of HydroBASINS, but has more fields. Data from several sources were used to annotate the river shapefile. The sources are listed in the documentation.

```{r read basinatlas , echo=TRUE, message=FALSE, warning=FALSE}
basinatlas <- "data/HydroATLAS_sample/sample__BasinATLAS_v10_lev10.shp" %>% 
  sf::st_read() %>% 
  sf::st_make_valid()
```


## HydroATLAS dataset

Both RiverATLAS and BasinATLAS have a lot of information (only the first 20 attributes are printed).

```{r check riveratlas and basinatlas , echo=TRUE, message=FALSE, warning=FALSE}
names(riveratlas)[1:20]
names(basinatlas)[1:20]
```

## HydroATLAS dataset

Digging a bit in the documentation: dis_m3_pyr is the average yearly discharge; for_pc_use is the percentage of forest cover in the upstream area; etc.

```{r discharge riveratlas forest basinatlas plot, echo=TRUE, message=FALSE, warning=FALSE}
plot_dis <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = riveratlas, aes(color = log10(dis_m3_pyr))) + 
  scale_color_viridis(name = "Discharge (m3/year)\nlog10-scale")

plot_forest <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) + 
  geom_sf(data = riveratlas, aes(color = for_pc_use)) + 
  scale_color_viridis(name = "Forest cover (%)")
```

## HydroATLAS dataset

```{r  discharge riveratlas plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_dis
```

## HydroATLAS dataset

```{r  forest basinatlas plot, echo=FALSE, message=FALSE, warning=FALSE}
plot_forest
```




# An example: Dams distribution in the Ebro basin

## Dams in the Ebro

The last step: we load a dataset of dams in the Ebro and we check the characteristics of the subcatchments where dams are located.

The dams data is retrieved from the [AMBER](https://amber.international/european-barrier-atlas/) European dataset. There are a lot of barriers, but we consider only dams (`dplyr::filter` function).

```{r  read dams, echo=TRUE, message=FALSE, warning=FALSE}
dams_ebro <- "data/dams_ebro/barriers_ebro.shp" %>% 
  st_read() %>% 
  dplyr::filter(LblAtls == "DAM")
```

## Dams in the Ebro

```{r  check dams, echo=TRUE, message=FALSE, warning=FALSE}
names(dams_ebro)
```

## Dams in the Ebro

A quick plot to check the distribution of dams in the Ebro basin. We can see that they are quite concentrated in the upper part of the basin, where the elevation is higher.

```{r  plot dams, echo=TRUE, message=FALSE, warning=FALSE}
plot <- ggplot() + 
  theme_bw() +
  geom_sf(data = hydrosheds_lev5, fill = NA, color = "gray50", alpha = 0.5) +
  geom_sf(data = hydrorivers, aes(alpha = ORD_STRA) ) +
  geom_sf(data = dams_ebro, color = "black", alpha = 0.5) +
  scale_alpha_continuous(name = "Stream order")

```

## Dams in the Ebro

```{r  plot dams plot, echo=FALSE, message=FALSE, warning=FALSE}
plot
```

## Dams in the Ebro

Next we check the characteristics of the subcatchments where dams are located. We use the function `sf::st_contains()` to check which subcatchments contain the dams. The function returns a list of indices of the dams contained by each subcatchment.

```{r  spatial join dams and basinatlas, echo=TRUE, message=FALSE, warning=FALSE}
dams_per_basin <- basinatlas %>% 
  sf::st_contains(dams_ebro)

dams_per_basin[[8]]
```

## Dams in the Ebro

Based on the dams_basin object, we count how many dams per subcatchment exist and we add this information to the basinatlas shapefile. Few subcatchments have a quite high number of dams.

```{r  count dams inn basinatlas, echo=TRUE, message=FALSE, warning=FALSE}
basinatlas <- basinatlas %>% 
  dplyr::mutate(
    n_dams = sapply(dams_per_basin, length))

plot_hist <- ggplot(basinatlas, aes(x = n_dams)) + 
  geom_histogram(binwidth = 1) +  theme_bw() + 
  xlab("Number of dams per subcatchment") + 
  ylab("Count of subcatchments")

```

## Dams in the Ebro

```{r  plot count dams inn basinatlas, echo=FALSE, message=FALSE, warning=FALSE}
plot_hist
```

## Dams in the Ebro

We can explore where the dams are located. For instance we can check if there is a relationship between the number of dams and elevation or slope. 

```{r  dams density relate, echo=TRUE, message=FALSE, warning=FALSE}
pl_dams <- basinatlas %>% 
  dplyr::select(n_dams, ele_mt_sav, slp_dg_sav, dis_m3_pyr) %>% 
  dplyr::filter(n_dams > 0) %>%
  pivot_longer(
    cols = c(ele_mt_sav, slp_dg_sav), 
    names_to = "variable",  values_to = "value") %>%
  ggplot() + theme_bw() +
  geom_point(aes(x = value, y = n_dams, color = log10(dis_m3_pyr))) +
  facet_wrap(~variable, scales = "free") + 
  scale_color_viridis(name = "Discharge (m3/year)\nlog10-scale")
```

## Dams in the Ebro

Many dams in high elevation and high slope areas, but also in areas with high discharge. Few big dams at low elevation.

```{r  dams density plot, echo=FALSE, message=FALSE, warning=FALSE}
pl_dams
```

# Thank you for listening!























