---
title: "Analysis of fragmented river systems"
subtitle: "Lecture 2: Generating river graphs"
author: "Damiano Baldan"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:
    widescreen: true
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, '2_river_networks.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## River systems as networks

-   River systems can be represented as graphs/networks
-   Nodes (vertices) represent river segments
-   Links (edges) represent the connections (confluences and barriers) between segments
-   Many connectivity indices are defined for graphs, straightforward application to river systems when represented as graphs
-   Inclusion of barriers complicates the representation, but several indices can be adapted to account for them (e.g. DCI, Dendritic Connectivity Index) 

## The topology of river networks

-   Planar networks: no links (edges) intersections
-   Tree-like networks: no loops, only one path between any two nodes (vertices)
-   Number of nodes (vertices) and links (edges) in a tree-like network: E = V - 1
-   Weighted networks: node weights can represent attributes of river segments (e.g. length, discharge, etc.), edge weights can represent the strength of the connection (e.g. presence/absence of a barrier, etc.)

## What we will cover in this presentation

-   The `igraph` package for graph manipulation in R
-   Processing of geographic data (e.g. HydroRIVERS, HydroBASINS, HydroATLAS datasets) to generate river networks without dams
-   processing of geographic data (e.g. AMBER) to generate river networks with dams
-   Most of the information is covered already in the [riverconn tutorial](https://damianobaldan.github.io/riverconn_tutorial/)

## Which packages we will use

-   `sf` and `terra` for vector and raster data manipulation

-   `dplyr` for data manipulation: helps building pipelines for easy-to-read data manipulation (loaded with the `tidyverse`)

-   `ggplot2` for data visualization: a bit painful, but easily integrates `sf` and `terra` objects for plotting (loaded with the `tidyverse`)

-   `igraph` for graph manipulation: useful for building/editing/managing graphs 


## Which packages we will use

```{r libraries, echo = TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(terra)
library(tidyterra)
library(viridis)
library(igraph)
```

# The `igraph` package

## The `igraph` package

-   The `igraph` package implements routines for graphs and network analysis. 
-   It can handle large graphs very well and provides functions for generating random and regular graphs, graph visualization, centrality methods and much more. 
-   The packages allows for easy construction of igraph objects based on symbolic description of edges and vertices, edges and vertices lists, or adjacency matrices. 
-   The book [Statistical Analysis of Network Data with R by Kolaczyk and Csardi (2014)](https://link.springer.com/book/10.1007/978-3-030-44129-6#toc) offers a comprehensive tutorial on the possibilities offered by the igraph package, including network data manipulation and visualization, descriptive analysis of network graphs, mathematical and statistical modeling of graphs, and network topology inference.

## The `igraph` package

We build a graph using a symbolic representation of vertices and edges. The notation 1--2 means that there is an edge between vertices 1 and 2.

```{r igraph toy graph 1, echo = TRUE}
g <- igraph::graph_from_literal(1--2, 2--5, 3--4, 4--5, 6--7)
plot(g)
```

## The `igraph` package

We check the graph. The graph has 7 vertices and 5 edged and is undirected (UN). The graph has the attribute name, that is a vertex attribute of class character. The graph has no edge attributes.

```{r igraph toy graph 1 check, echo = TRUE}
g
```

## The `igraph` package

We build a directed graph using a symbolic representation of vertices and edges. The notation 1-+2 means that there is an edge with direction from 1 and 2. 

```{r igraph  toy graph 2, echo = TRUE}
g <- igraph::graph_from_literal(1-+2, 2-+5, 3-+4, 4-+5, 6-+7)
plot(g)
```

## The `igraph` package

We check the graph. The graph has 7 vertices and 5 edged and is directed (DN). The graph has the attribute name, that is a vertex attribute of class character. The graph has no edge attributes.

```{r igraph toy graph 2 check, echo = TRUE}
g
```

## The `igraph` package

We create a graph that is a bit more complex. 

```{r igraph toy graph 3, echo = TRUE}
g <- igraph::graph_from_literal( 1-+2, 2-+5, 3-+4, 4-+5, 6-+7,
  7-+10, 8-+9, 9-+10, 5-+11, 11-+12, 10-+13, 13-+12, 12-+14, 14-+15, 15-+16)
plot(g)
```

## The `igraph` package

We can transform this graph to an adjacenty matrix (`as_adjacency_matrix()`; 1 = there is an edge). Note that the matrix is not symmetric because the graph is directed.  

```{r igraph toy graph 3 as matrix, echo = TRUE, message=FALSE, warning=FALSE}
mat <- igraph::as_adjacency_matrix(g)
mat[1:8, 1:8]
```

## The `igraph` package

The graph can be generated back from the adjacency matrix (`graph_from_adjacency_matrix()`). We check if the new graph is isomorphic (`isomorphic` function) to the original one (i.e. they are the same graph, but with different vertex labels).  

```{r igraph toy graph 3 from matrix, echo = TRUE, message=FALSE, warning=FALSE}
g1 <- igraph::graph_from_adjacency_matrix(mat)
igraph::isomorphic(g1, g)
```

## The `igraph` package

We can transform this graph to a data.frame (`as_data_frame()`). Note that the data.frame has two columns, `from` and `to`, that represent the source and target vertices of each edge.   

```{r igraph toy graph 3 as df, echo = TRUE, message=FALSE, warning=FALSE}
g_df <- igraph::as_data_frame(g, "edges")
head(g_df)
```

## The `igraph` package

The graph can be generated back from the data.frame (`graph_from_data_frame()`). We check if the new graph is isomorphic (`isomorphic`) to the original one (i.e. they are the same graph, but with different vertex labels).  

```{r igraph toy graph 3 from df, echo = TRUE, message=FALSE, warning=FALSE}
g1 <- igraph::graph_from_data_frame(g_df)
igraph::isomorphic(g1, g)
```

# Centrality metrics with the `igraph` package

## Centrality metrics with the `igraph` package

-   Centrality metrics are used to identify the most important vertices in a graph
-   Identify nodes that are relevant for connecting separate sections of the graph
-   Degree centrality: number of edges connected to a vertex (for weighted networks: strenght)
-   Betweenness centrality: number of shortest paths that pass through a vertex
-   Closeness (harmonic) centrality: inverse of the average shortest path length from a vertex
-   For directed and undirected graphs

## Centrality metrics with the `igraph` package

Degree centrality of node $i$ ($k_{i}$): number of edges connected to a vertex (for weighted networks: strenght)

$$k_i = \sum_{j}a_{ij}$$
where $a_{ij}$ is the element of the adjacency matrix that represents the presence (1) or absence (0) of an edge between vertices $i$ and $j$

## Centrality metrics with the `igraph` package

Betweenness centrality: number of shortest paths that pass through a vertex

$$g_i = \sum_{s \neq i \neq t} \frac{\sigma_{st}(i)}{\sigma_{st}}  $$
where where $\sigma_{st}$ is the total number of shortest paths from node $s$ to node $t$ and $\sigma_{st}(i)$ is the number of those paths that pass through node $i$ (not where $i$ is an end point)

## Centrality metrics with the `igraph` package

Closeness centrality: inverse of the average shortest path length from a vertex

$$c_i = \frac{1}{\sum_{j \neq i} {d(ij)}}  $$
Harmonic centrality: for disconnected graphs (to avoid issues with $d(ij) = \infty$)

$$c_i = \sum_{j \neq i} \frac{1}{ {d(ij)}}  $$

where $d(ij)$ is the shortest distance between nodes $i$ and $j$


## Centrality metrics with the `igraph` package

We create a graph, calculate degree, betweeness and closeness centrality and store the values as nodes attributes

```{r igraph centrality calculation, echo = TRUE, message=FALSE, warning=FALSE}
g <- igraph::graph_from_literal( 1-+2, 2-+5, 3-+4, 4-+5, 6-+7,
  7-+10, 8-+9, 9-+10, 5-+11, 11-+12, 10-+13, 13-+12, 12-+14, 14-+15, 15-+16)
V(g)$degree <- igraph::degree(g)
V(g)$betweenness <- igraph::betweenness(g)
V(g)$closeness <- igraph::closeness(g, mode = "all")
g
```
## Centrality metrics with the `igraph` package

We plot the graph with vertex color proportional to the centrality metrics

```{r igraph centrality plot, echo = TRUE, message=FALSE, warning=FALSE}
library(ggnetwork)
gg0 <- ggnetwork(g)

pl1 <- ggplot(gg0, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(arrow = arrow(length = unit(0.5, "lines"), type = "closed")) + 
  geom_nodes(aes(color = degree), size = 5) + theme_void()

pl2 <- ggplot(gg0, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(arrow = arrow(length = unit(0.5, "lines"), type = "closed")) + 
  geom_nodes(aes(color = closeness), size = 5) + theme_void()

pl3 <- ggplot(gg0, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(arrow = arrow(length = unit(0.5, "lines"), type = "closed")) + 
  geom_nodes(aes(color = betweenness), size = 5) + theme_void()

plot <- ggpubr::ggarrange(pl1, pl2, pl3, ncol = 3, nrow = 1)

```
## Centrality metrics with the `igraph` package

```{r igraph centrality plot plot, echo = FALSE}
plot
```


# The Ebro case study

## The Ebro case study

We load the input data in the environment. We have three shapefiles: one for the river network, one for the river basin and one for the dams. The dams are from a local dataset, that contains also informations on the upstream reservoir.

```{r river data input read, echo = TRUE}
shape_river <- "data/HydroRIVERS_sample/sample__hydroRIVERS_v10_eu.shp" %>% 
  st_read(quiet = TRUE)
shape_basin <- "data/HydroBASINS_sample/sample__hybas_eu_lev10_v1c.shp" %>% 
  st_read(quiet = TRUE)
shape_dams <- "data/dams_ebro/embalses.shp" %>% 
  st_read(quiet = TRUE)
```

Then, we load the content of the file `functions_network_creation.R`, that contains several useful functions.

```{r utility funs source, echo = TRUE}
source("functions_network_creation.R")
```

## The Ebro case study

The river shapefile has a very high resolution. To speed up the analyses, we drop all the river segments that have an upland area greater than 80 km2.

```{r prune river network, echo = TRUE}
shape_river_small <- shape_river %>% 
  filter(UPLAND_SKM > 80)

nrow(shape_river)
nrow(shape_river_small)

```

## The Ebro case study

We convert the dams shapefile from polygons to points (the polygons are the borders of the reservoir).

```{r plot everything, echo = TRUE, message = FALSE}
dams_to_points <- shape_dams %>%
  st_as_sf() %>% st_centroid() %>%
  mutate(id = 1:nrow(.)) %>%  dplyr::select(id) %>% 
  st_transform(crs = "+proj=longlat +datum=WGS84")

plot <- ggplot() +
  coord_fixed() +
  theme_minimal() +
  ggspatial::layer_spatial(shape_basin, fill = "white", color = "gray90") +
  ggspatial::layer_spatial(shape_river_small, aes(color = log10(UPLAND_SKM)) )+
  ggspatial::layer_spatial(dams_to_points) +
  scale_color_viridis(direction = -1, name= "upstream area \n(log10[Km^2])") +
  theme(legend.position = "bottom") +
  labs(caption = "Black dots are the position of the dams")

```

## The Ebro case study

```{r plot everything , echo = FALSE, message = FALSE}
plot
```

# Hydrographic data processing

## Hydrographic data processing

The first step before building the graph is to identify confluences. To do this, the river network is split into the smallest segments allowed by its geometry.

```{r identify confluences 1, echo = TRUE, message = FALSE}
shape_river_simple <- shape_river_small %>%
  st_as_sf %>%
  st_union()

river_to_points <- shape_river_simple  %>%
  st_as_sf() %>%
  st_cast("POINT") %>%
  mutate(id = 1:nrow(.))

```

## Hydrographic data processing

Confluences are the points where three or more segments overlap.

```{r identify confluences 2, echo = TRUE, message = FALSE}
joins_selection <- river_to_points %>%
  st_equals() %>%
  Filter(function(x){length(x) > 2}, .) %>%
  lapply(., FUN = min) %>%
  unlist() %>%
  unique()

river_joins <- river_to_points %>% 
  filter(id %in% joins_selection)

```

## Hydrographic data processing

Finally, the simplified shapefile is split where conferences are.

```{r identify confluences 3, echo = TRUE, message = FALSE}
shape_river_simplified <- lwgeom::st_split(shape_river_simple, river_joins) %>%
  st_collection_extract(.,"LINESTRING") %>%
  data.frame(id = 1:length(.), geometry = .) %>%
  st_as_sf() %>%
  mutate(length = st_length(.))

```

## Hydrographic data processing

We plot to make sure everything was succesful.

```{r identify confluences 4, echo = TRUE, message = FALSE}
plot <- ggplot() +
  coord_fixed() +
  ggspatial::layer_spatial(shape_river_simplified, aes(color = id))+
  scale_color_viridis(direction = -1, name= "Reach ID") +
  ggspatial::layer_spatial(river_joins, shape = 1)+
  theme_minimal() +
  theme(legend.position = "bottom")+
  labs(caption = "Hollow points are the position of the junctions")

```

## Hydrographic data processing

```{r identify confluences 4 plot, echo = FALSE, message = FALSE}
plot
```


# Barriers data processing

## Barriers data processing

We use the custom `snap_to_river()` function to overlay dams with the river network.

```{r dams processing 1, echo = TRUE, message = FALSE, warning = FALSE}
dams_snapped <- snap_to_river(dams_to_points,
                              shape_river_simple %>% st_sf(),
                              max_dist = 1000)

dams_snapped_reduced <-
  dams_snapped[
    st_contains(shape_river_simple %>% st_sf(), 
                dams_snapped, prepared = FALSE)[[1]],]

st_distance(dams_snapped_reduced, shape_river_simple) %>% sum

```

## Barriers data processing

If two or more dams are snapped to the same point, they are considered as one dam. This does not work with 

```{r dams processing 2, echo = TRUE, message = FALSE, warning = FALSE}
dams_snapped_reduced_joined <- dams_snapped_reduced %>%
  mutate(cluster =
           st_equals(dams_snapped_reduced, dams_snapped_reduced) %>%
           sapply(., FUN = function(x) paste0(x, collapse = "_"))) %>%
  group_by(cluster) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(id_dam = as.character(1:nrow(.)), pass_u = 0.1, pass_d = 0.8) %>%
  as.data.frame %>%
  st_as_sf() %>%
  st_join(., shape_river_small, join = st_is_within_distance, dist = 10 ) %>% 
  group_by(id) %>%
  slice(which.max(UPLAND_SKM)) %>% 
  ungroup()

```

## Barriers data processing

The custom function `headwaters_dam()` is useful for identifying dams that do not have an upstream reach and would not be contributing to fragmentation.

```{r dams processing 3, echo = TRUE, message = FALSE, warning = FALSE}
headwaters_checking <- headwaters_dam(dams_snapped_reduced_joined, shape_river_simple)

headwaters_checking$flag_headwater %>% sum()

```


## Barriers data processing

Now dams can be plotted. 

```{r dams plot, echo = TRUE, message = FALSE, warning = FALSE}
plot <- ggplot() +
  coord_fixed() +
  ggspatial::layer_spatial(shape_river_simple, color = "gray70")+
  ggspatial::layer_spatial(dams_snapped_reduced, shape = 1) +
  theme_minimal() +
  theme(legend.position = "bottom")+
  ggspatial::annotation_north_arrow(location = "br")+
  ggspatial::annotation_scale(location = "bl", style = "ticks") +
  labs(caption = "Hollow points are the position of the dams")

```

## Barriers data processing

```{r dams plot plot, echo = FALSE, message = FALSE, warning = FALSE}
plot 
```

# Integrating hydrographic and dams data

## Integrating hydrographic and dams data

The position of confluences and dams are integrated in the same data.frame, that will be used to build the graph. The type of link (confluence or dam) is stored in the column `type`, while the columns `pass_u` and `pass_d` store the passability values for upstream and downstream movement, respectively.

```{r dams confluences info merge, echo = TRUE, message = FALSE, warning = FALSE}
network_links <- rbind(
  dams_snapped_reduced_joined %>% 
    mutate(type = "dam", id_barrier = id_dam) %>%
    dplyr::select(type, id_barrier, pass_u, pass_d),
  river_joins %>% mutate(type = "joint") %>%
    dplyr::select(type) %>%
    mutate(id_barrier = NA, pass_u = NA, pass_d = NA) %>%
    rename(geometry = x)) %>%
  mutate(id_links = 1:nrow(.))

```

## Integrating hydrographic and dams data

Finally, the shapefile is sliced based on confluences and dams positions.

```{r final river shapefile slice, echo = TRUE, message = FALSE, warning = FALSE}
river_net_simplified <- lwgeom::st_split(shape_river_simple, network_links) %>%
  st_collection_extract(.,"LINESTRING") %>%
  data.frame(NodeID = 1:length(.), geometry = .) %>%
  st_as_sf() %>%
  mutate(length = st_length(.)) %>%
  st_join(., shape_river_small, join = st_is_within_distance, dist = 0.01 ) %>% 
  group_by(NodeID) %>%
  slice(which.max(UPLAND_SKM)) %>% 
  ungroup()

```

## Integrating hydrographic and dams data

We check with a plot

```{r final river shapefile sliced plot, echo = TRUE, message = FALSE, warning = FALSE}
plot <- ggplot() +
  coord_fixed() +
  ggspatial::layer_spatial(river_net_simplified, color = "gray70")+
  ggspatial::layer_spatial(network_links, aes(shape = type))+
  scale_shape_manual(name = "Splitting points", values=c("dam" =17,"joint" = 23))+
  theme_minimal() +
  ggspatial::annotation_scale(location = "bl", style = "ticks")

```

## Integrating hydrographic and dams data

We check with a plot

```{r final river shapefile sliced plot plot, echo = FALSE, message = FALSE, warning = FALSE}
plot
```


# Troubleshooting

## Troubleshooting

A common issue when generating the river network is when more than three segments are overlapping in a confluence. The custom function `multiple_confluences()` detects such issues. The function returns a shapefile that can be saved and opened in QGIS to find and correct manually the issue.

```{r multiple confluences, echo = TRUE, message = FALSE, warning = FALSE}
confluences <- multiple_confluences(river_net_simplified)
sum(confluences$flag_confluence)
```

## Troubleshooting

A common issue when generating the river network is when more than three segments are overlapping in a confluence. The custom function `multiple_confluences()` detects such issues. The function returns a shapefile that can be saved and opened in QGIS to find and correct manually the issue.

```{r shp check, echo = TRUE, message = FALSE, warning = FALSE}
shp_check <- check_components(network_links, river_net_simplified)
head(shp_check)
```


























