---
title: "Analysis of fragmented river systems"
subtitle: "Lecture 4: Assessing fragmentation impacts on riverine metacommunities"
author: "Damiano Baldan"
date: "2026-02-25"
output: 
  ioslides_presentation:
    widescreen: true
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, '4_fragmented_metacomm.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## What we will cover in this presentation

-   Why fragmentation is important for riverine metacommunities (alpha and beta diversity)
-   Example of application of `riverconn` indices to fish metacommunities in the Ebro basin
-   Generalized linear models for alpha diversity
-   Mantel test and multilinear regression for beta diversity
-   Generalized linear models for the local contribution to beta diversity (LCBD)


## Fragmentation and metacommunities

-   Metacommunity: a set of local communities linked by dispersal
-   Local communities are influenced by both local environmental conditions and regional dispersal processes
-   Mass effect (ME) archetype: high dispersal -> 'everything is everywhere'
-   Species sorting (SS) archetype: low dispersal -> high effect of local environment (filter)
-   Dispersal (connectivity) strength regulates between ME and SS


## Fragmentation and metacommunities

-   In a landscape, central sites are more connected and peripherical sites are more isolated
-   We expect effect of centrality/isolation on alpha and beta diversity
-   River network have per se strong spatial structure
-   dams affect this spatial structure
-   `riverconn` indices can help quantifying this 'new' spatial structure


## Which packages we will use

-   `dplyr` for data manipulation: helps building pipelines for easy-to-read data manipulation (loaded with the `tidyverse`)
-   `ggplot2` for data visualization: a bit painful, but easily integrates `sf` and `terra` objects for plotting (loaded with the `tidyverse`)
-   `vegan` for ecological analyses: provides functions for calculating diversity indices and performing Mantel tests
-   `MASS` for the negative binomial GLM 
-   `adespatial` for calculating beta diversity
-   `ecodist` for the Mantel test with multiple predictors
-   `betareg` and `StepBeta` for Generalized Linear Models with beta link function


## Which packages we will use

```{r libraries, echo = TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(viridis)
library(igraph)
library(riverconn)
library(vegan)
library(MASS)
library(adespatial)
library(ecodist)
library(betareg)
library(StepBeta)
```

# The Ebro case study: input data

## The Ebro case study: input data

We use fish metacommunities data collected in the Ebro. The data were already processed. First we read all the input data, then we check them.

```{r input data read , echo = TRUE, message=FALSE, warning=FALSE}
species_mat_pa  <- read.table("data/metacomm_data/ebro_species_mat_pa.txt")

env_mat <- read.table("data/metacomm_data/ebro_env_mat.txt") %>% 
  dplyr::mutate( across( where(is.character), factor)) 

frag_mat <- read.table("data/metacomm_data/ebro_frag_mat.txt")

centr_mat <- read.table("data/metacomm_data/ebro_centr_mat.txt")

dist_mat <- readRDS("data/metacomm_data/dist_mat_ebro.RDS") 
dist_mat <- dist_mat[rownames(species_mat_pa), rownames(species_mat_pa)]

# # species_mat_inv_pa  <- read.table("data/metacomm_data/ebro_species_mat_inv_pa.txt")
# frag_mat <- read.table("data/metacomm_data/ebro_frag_mat.txt")
# # frag_mat_c <- read.table("data/metacomm_data/ebro_frag_mat_c.txt")
# beta <- readRDS("data/metacomm_data/ebro_beta_dist_mat.RDS")
# centr_mat <- read.table("data/metacomm_data/ebro_centr_mat.txt")
# LCBD_df <- read.table("data/metacomm_data/ebro_LCBD_df.txt")
# dist_mat <- readRDS("data/metacomm_data/dist_mat_ebro.RDS")
```


## The Ebro case study: input data

We check the species-sites matrix (`species_mat_pa`). There's not many species in the Ebro, and most of the sites have few species.

```{r species data check 1, echo = TRUE, message=FALSE, warning=FALSE}
nrow(species_mat_pa)
ncol(species_mat_pa)
```

## The Ebro case study: input data

We check the species-sites matrix (`species_mat_pa`). There's not many species in the Ebro, and most of the sites have few species.

```{r species data check 2, echo = TRUE, message=FALSE, warning=FALSE}
names(species_mat_pa)
rowSums(species_mat_pa) %>% summary()
```

## The Ebro case study: input data

We read the sampling points shapefile and plot the species richness.

```{r species data check 3, echo = TRUE, message=FALSE, warning=FALSE}
richness <- rowSums(species_mat_pa) %>% 
  data.frame(Site_ID = names(.), richness = .)

sampling_points <- st_read("data/metacomm_data/fish_Ebro_points.shp", quiet = TRUE) %>% 
  dplyr::filter(Site_ID %in% rownames(species_mat_pa) ) %>% 
  left_join(richness, by = "Site_ID")

river_shapefile <- st_read("data/river_graph/river_net_simplified.shp", quiet = TRUE)
```

## The Ebro case study: input data

We read the sampling points shapefile and plot the species richness.

```{r species data check 3 plot, echo = TRUE, message=FALSE, warning=FALSE}
plot <- ggplot() +  theme_minimal() +
  geom_sf(data = river_shapefile, color = "gray50", size = 0.2) +
  geom_sf(data = sampling_points, aes(color = richness), alpha = 0.5) +
  scale_color_viridis(name = "Species richness") +
  theme(legend.position = "right")
```
## The Ebro case study: input data

We read the sampling points shapefile and plot the species richness.

```{r species data check 3 plot plot, echo = TRUE, message=FALSE, warning=FALSE}
plot
```

## The Ebro case study: input data

We check the environmental data matrix (`env_mat`). We have 11 covariates.

```{r env data check, echo = TRUE, message=FALSE, warning=FALSE}
nrow(env_mat)
ncol(env_mat)
names(env_mat)
```

## The Ebro case study: input data

From the fragmentation matrix (`frag_mat`) we consider the DCI values calculated with: uniform habitat weight, symmetric dispersal (`param = c(0.3, 0.6, 0.9)`), and `pass_u = pass_d = 0.5` 

```{r frag data check, echo = TRUE, message=FALSE, warning=FALSE}
frag_mat %>% 
  dplyr::select(index_us0.3,index_us0.6,index_us0.9 ) %>% 
  summary()
```

## The Ebro case study: input data

We consider also two classic network centrality metrics: betweenness centrality and out-closeness centrality. 

```{r centr data check, echo = TRUE, message=FALSE, warning=FALSE}
centr_mat %>% 
  dplyr::select(BCdir, CCout) %>% 
  summary()
```
## The Ebro case study: input data

Finally, we also check the distance matrix (distances in meters). 

```{r dist mat data check, echo = TRUE, message=FALSE, warning=FALSE}
hist(dist_mat)
```
## The Ebro case study: input data

We have all the data we need. We put all the predictors in a single data.frame. We logtransform fragmentation variables. We logit-transform land use variables. Finally, we scale all the variables. We drop a couple of variables that are highli correlated. 

```{r assembling predictors, echo = TRUE, message=FALSE, warning=FALSE}
pred_mat_lm_full <- cbind(
  env_mat %>% dplyr::select(elev, Dischrg) %>%  rename(size = Dischrg) %>%  dplyr::mutate(size = (size)^0.25),
  env_mat %>% dplyr::select(AGR, UR, FR) %>% car::logit(),
  frag_mat %>% dplyr::select(index_us0.3, index_us0.6, index_us0.9) %>%
    dplyr::mutate(index_us0.3 = log10(index_us0.3),
                  index_us0.6 = log10(index_us0.6),
                  index_us0.9 = log10(index_us0.9)) %>%
    rename_all(~stringr::str_replace(.,"index_","")),
  centr_mat %>% dplyr::select(BCdir, CCout) ) %>%
  mutate_if(is.numeric, ~(scale(.) %>% as.vector))

cor_mat <- pred_mat_lm_full %>%  cor(method = "spearman")
pred_mat_lm <- pred_mat_lm_full %>% dplyr::select(-us0.6, -FR)

```


## The Ebro case study: input data

```{r plotting predictors, echo = TRUE, message=FALSE, warning=FALSE}
corrplot::corrplot(cor_mat, type = "lower", method = "color", tl.col = "black", tl.srt = 45,
    addCoef.col = "black", number.cex = 0.7, diag = FALSE)

```

# The Ebro case study: alpha diversity

## The Ebro case study: alpha diversity

We want to analyze the drivers of alpha diversity. We use species richness and we employ a Generalized Linear Model with a negative binomial link function. We use a forward selection to select a parsimonious model. 

```{r alpha model, echo = TRUE, message=FALSE, warning=FALSE}
rich <- rowSums(species_mat_pa)
```

## The Ebro case study: alpha diversity

```{r alpha model 2, echo = TRUE, message=FALSE, warning=FALSE}
hist(rich)
```

## The Ebro case study: alpha diversity

We want to analyze the drivers of alpha diversity. We use species richness and we employ a Generalized Linear Model with a negative binomial link function. We use a forward selection to select a parsimonious model. 

```{r alpha model step, echo = TRUE, message=FALSE, warning=FALSE}
lm_rich <- step(
  object = glm.nb(rich ~ 1, data = pred_mat_lm),
  scope = formula(glm.nb(rich ~ ., data = pred_mat_lm)),
  direction = "forward", trace = 0)
```

## The Ebro case study: alpha diversity

```{r alpha model summary, echo = TRUE, message=FALSE, warning=FALSE}
summary(lm_rich)
```

## The Ebro case study: alpha diversity

We see that fragmentation has some effect, but not the way we intend it.

```{r alpha model plotmo,  echo=FALSE, results='hide',fig.keep='all'}
plotmo::plotmo(lm_rich)
```

# The Ebro case study: beta diversity

## The Ebro case study: beta diversity

Next we want to analyze what are the drivers of beta diversity. First we calculate the beta diversity matrix with `adespatial::beta.div()` function. We use the Sorensen indices, and we are interested in three components of the beta diversity: the total beta diversity (D), the richness difference (rich), and the replacement (repl), as they capture different community asssembly processes.

```{r beta calculation,  echo=FALSE, results='hide',fig.keep='all'}
beta <- beta.div.comp(species_mat_pa, coef = "S")
names(beta)
```
## The Ebro case study: beta diversity

We use a Mantel test (`ecodist::mantel()`) to check the influence of fragmentation on beta diversity. First we calculate distances between predictors. We also retain the lower triangle of the distance matrix for later use.

```{r beta predictors,  echo = TRUE, message=FALSE, warning=FALSE}
dist_data <- cbind(
  "AGR" = ecodist::distance(env_mat$AGR  %>% car::logit() %>% scale %>% as.vector),
  "UR" = ecodist::distance(env_mat$UR  %>% car::logit() %>% scale %>% as.vector),
  "us0.3" = ecodist::distance(frag_mat$index_us0.3 %>% scale %>% as.vector),
  "us0.6" = ecodist::distance(frag_mat$index_us0.6 %>% scale %>% as.vector),
  "us0.9" = ecodist::distance(frag_mat$index_us0.9 %>% scale %>% as.vector),
  "elev" = ecodist::distance(env_mat$elev %>% scale %>% as.vector),
  "size" = ecodist::distance(env_mat$Dischrg ^ 0.25 %>% scale %>% as.vector),
  "bc" = ecodist::distance(centr_mat$BCdir %>% scale %>% as.vector),
  "cc" = ecodist::distance(centr_mat$CCout %>% scale %>% as.vector)
) %>% as.data.frame()

dist_mat_v <- dist_mat[lower.tri(dist_mat)]
```

## The Ebro case study: beta diversity

Next, we use the Mantel test check the influence of fragmentation on beta diversity. We partial out the effect of distance. The tests are marginally significant.

```{r beta mantel 1,  echo = TRUE, message=FALSE, warning=FALSE}
ecodist::mantel(as.vector(beta$D) ~ dist_data$us0.9 + dist_mat_v, mrank = TRUE)
ecodist::mantel(as.vector(beta$D) ~ dist_data$us0.3 + dist_mat_v, mrank = TRUE)
```
## The Ebro case study: beta diversity

Instead we check the effect of elevation and size, and they are both highly significant.

```{r beta mantel 2,  echo = TRUE, message=FALSE, warning=FALSE}
ecodist::mantel(as.vector(beta$D) ~ dist_data$elev + dist_mat_v, mrank = TRUE)
ecodist::mantel(as.vector(beta$D) ~ dist_data$size + dist_mat_v, mrank = TRUE)
```
## The Ebro case study: beta diversity

We check the effect of fragmentation on the replacement component.

```{r beta mantel 3,  echo = TRUE, message=FALSE, warning=FALSE}
ecodist::mantel(as.vector(beta$repl) ~ dist_data$us0.9 + dist_mat_v, mrank = TRUE)
ecodist::mantel(as.vector(beta$repl) ~ dist_data$us0.3 + dist_mat_v, mrank = TRUE)
```

## The Ebro case study: beta diversity

We check the effect of fragmentation on the richness difference component.

```{r beta mantel 4,  echo = TRUE, message=FALSE, warning=FALSE}
ecodist::mantel(as.vector(beta$rich) ~ dist_data$us0.9 + dist_mat_v, mrank = TRUE)
ecodist::mantel(as.vector(beta$rich) ~ dist_data$us0.3 + dist_mat_v, mrank = TRUE)
```

## The Ebro case study: beta diversity

To facilitate the comparison we can use the multilinear regression on the distance matrix, that allows for multiple predictors. No significant effect of fragmentation.

```{r beta mrm beta 1,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model <- MRM(formula = as.vector(beta$D) ~., data = dist_data, mrank = TRUE)
mrm_model$coef
```
## The Ebro case study: beta diversity

To facilitate the comparison we can use the multilinear regression on the distance matrix, that allows for multiple predictors.

```{r beta mrm beta 2,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model$r.squared
mrm_model$F.test
```

## The Ebro case study: beta diversity

We fit a MRM model to the replacement component. Marginally significant effect of fragmentation. 

```{r beta mrm repl 1,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model <- MRM(formula = as.vector(beta$repl) ~., data = dist_data, mrank = TRUE)
mrm_model$coef
```
## The Ebro case study: beta diversity

We fit a MRM model to the replacement component. Significant effect of fragmentation.

```{r beta mrm repl 2,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model$r.squared
mrm_model$F.test
```

## The Ebro case study: beta diversity

We fit a MRM model to the richness difference component. Significant effect of fragmentation.

```{r beta mrm rich 1,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model <- MRM(formula = as.vector(beta$rich) ~., data = dist_data, mrank = TRUE)
mrm_model$coef
```
## The Ebro case study: beta diversity

We fit a MRM model to the richness difference component. Marginally significant effect of fragmentation.

```{r beta mrm rich 2,  echo = TRUE, message=FALSE, warning=FALSE}
mrm_model$r.squared
mrm_model$F.test
```


# The Ebro case study: Local Contribution to Beta Diversity (LCBD)

## The Ebro case study: LCBD

Finally, we analyze LCBD, that is the contribution of each site to the total beta diversity. We check if LCBD is related to fragmentation. We first calculate LCBDs.

```{r beta lcbd,  echo = TRUE, message=FALSE, warning=FALSE}
LCBD_df <- data.frame(
  "site" = rownames(species_mat_pa),
  "LCBD_repl" = LCBD.comp(beta[[1]] ^ 0.5 )[[2]],
  "LCBD_rich" = LCBD.comp(beta[[2]] ^ 0.5 )[[2]],
  "LCBD_D" = LCBD.comp(beta[[3]] ^ 0.5 )[[2]]  ) %>%
  column_to_rownames("site") %>%
  mutate(LCBD_repl = ifelse(LCBD_repl <0, abs(LCBD_repl), LCBD_repl))
```

## The Ebro case study: LCBD

We fit a Generalized Linear Model with a beta link function.

```{r beta lcbd beta,  echo = TRUE, message=FALSE, warning=FALSE, results='hide'}
beta <- LCBD_df$LCBD_D
start <- betareg::betareg(beta ~ ., data = pred_mat_lm)
breg <- StepBeta::StepBeta(start)
breg$pseudo.r.squared
```
## The Ebro case study: LCBD

```{r beta lcbd beta summary,  echo = FALSE, message=FALSE, warning=FALSE}
summary(breg, FALSE)
```
## The Ebro case study: LCBD

We fit a Generalized Linear Model with a beta link function. We fit the replacement LCBDs.

```{r beta lcbd repl,  echo = TRUE, message=FALSE, warning=FALSE, results='hide'}
beta <- LCBD_df$LCBD_repl
start <- betareg::betareg(beta ~ ., data = pred_mat_lm)
breg <- StepBeta::StepBeta(start)
breg$pseudo.r.squared
```
## The Ebro case study: LCBD

```{r beta lcbd repl summary,  echo = FALSE, message=FALSE, warning=FALSE}
summary(breg, FALSE)
```

## The Ebro case study: LCBD

We fit a Generalized Linear Model with a beta link function. We fit the richness difference LCBDs.

```{r beta lcbd rich,  echo = TRUE, message=FALSE, warning=FALSE, results='hide'}
beta <- LCBD_df$LCBD_rich
start <- betareg::betareg(beta ~ ., data = pred_mat_lm)
breg <- StepBeta::StepBeta(start)
breg$pseudo.r.squared
```
## The Ebro case study: LCBD

```{r beta lcbd rich summary,  echo = FALSE, message=FALSE, warning=FALSE}
summary(breg, FALSE)
```


# Thank you for listening!























